---
title: "EchoGO interpretation and hypothesis follow-up"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{EchoGO interpretation and hypothesis follow-up}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# The v0.1.4 evidence model

EchoGO is a traceable functional-interpretation and hypothesis-generation
framework. It preserves the target GOseq result, evaluates the same experiment
through researcher-selected annotation contexts, and records the source,
context, identifier, recognition, and available annotation provenance for every
reported functional interpretation.

It helps answer three separate questions:

1. What does the target analysis already show?
2. Which selected annotation contexts recover the same exact functional term?
3. What additional, traceable functions should be investigated?

Selected organisms are annotation contexts, not independent experimental
replicates and not automatic evidence of conservation.

## Read the primary evidence table

`consensus/term_evidence_exact.csv` is the canonical exact-term table. It does
not contain a composite score or an aggregated p-value. Its key fields are:

- `target_goseq_supported`: whether the term passes the configured GOseq
  source-specific adjusted-p threshold;
- `gprof_custom_context_support_n` and
  `alternative_context_support_n`: the number of qualifying custom-background
  g:Profiler contexts, and the subset that are alternative contexts;
- `alternative_context_support_fraction`: recurrence across the declared
  researcher-selected alternatives;
- `evidence_profile`: one of the primary classes below; and
- `source_provenance_references`: links to the long-form source rows.

Source-specific adjusted p-values and fold enrichment remain in
`consensus/term_source_provenance_long.csv`. They must be interpreted within
their own GOseq or g:Profiler analysis; they are never minimized, maximized, or
combined across contexts.

## Primary evidence profiles

| Profile | Meaning |
|---|---|
| `TARGET_ONLY` | A qualifying target GOseq term without qualifying recovery in an alternative custom-background context. |
| `TARGET_PLUS_CONTEXT` | A qualifying target GOseq term recovered in one or more alternative custom-background contexts. This describes cross-context representation of the exact term. |
| `ALTERNATIVE_CONTEXT` | Not a qualifying target GOseq term, but recovered in one or more alternative custom-background contexts. This is a traceable hypothesis for follow-up. |

The primary profiles never include default-domain evidence. A row with only
default-domain evidence is retained as `NO_PRIMARY_SUPPORT`, not recast as a
primary finding.

## Target and alternative contexts

The researcher supplies the full requested context panel and an explicit
target-context declaration. If `target_context` is set (for
example, `"drerio"` in a zebrafish analysis), it is labelled `TARGET` and is
excluded from alternative-context recurrence. If it is `NA_character_` (for
example, a Gammarus analysis with no target g:Profiler organism), all requested
contexts are alternatives. Target GOseq remains the experimental anchor.

EchoGO does not infer these roles from phylogeny and does not construct
per-species ortholog vectors. The same resolved portable foreground and
experimental background are submitted to each selected context. Context-specific
effective query/domain sizes are recognition and statistical-domain metadata,
not ortholog counts.

## Follow up an `ALTERNATIVE_CONTEXT` term

Use `echogo_hypothesis_drilldown()` with the exact-term table and its long-form
source provenance. The returned object contains the term row, each source row,
the contributing experimental genes, and matching annotation provenance when it
is available.

```{r eval=FALSE}
drilldown <- echogo_hypothesis_drilldown(
  evidence_table = exact_terms,
  source_provenance = source_provenance,
  term_id = "GO:0000001",
  annotation_provenance = gene_provenance
)
```

This evidence trail supports a biological follow-up decision. It does not prove
a mechanism, independent replication, conservation, or a new combined
significance statistic.

## Optional default-domain exploration

Set `run_exploratory_default_domain = TRUE` only when the broader
g:Profiler default annotated domain is intentionally wanted. Its canonical
label is `default_domain_exploratory`; the primary matched-background label is
`custom_experimental_background`.

Default-domain results are a separate optional exploratory layer. They cannot
change primary profiles, primary recurrence, primary plots, semantic reduction,
networks, representative selection, or evaluation.

## Annotation provenance

When supplied by the annotation input, EchoGO writes:

- `annotation_provenance_gene_level.csv`;
- `annotation_provenance_summary.csv`; and
- `term_annotation_provenance_summary.csv`.

The resolver always records original identifiers, resolved portable names and
resolution routes. It additionally preserves available native symbols,
SwissProt fields, eggNOG preferred names, taxonomy, annotation-match status,
and supported richer eggNOG fields. Missing provenance remains missing. These
fields are descriptive and never form a confidence score.
