---
title: "EchoGO workflow: traceable multi-context interpretation"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{EchoGO workflow: traceable multi-context interpretation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Inputs and analysis boundary

EchoGO accepts an annotated GOseq result, a DE table or an explicit tested
universe, and an annotation table such as a Trinotate/eggNOG export. The
experimental tested universe is never inferred from GO terms. The significant
foreground is selected by the declared DE rule and must be a subset of the
tested background.

The resolver constructs one portable foreground and one portable background
from those experimental vectors. It submits the same resolved vectors to every
researcher-selected g:Profiler context. It does not create organism-specific
ortholog vectors.

```{r eval=FALSE}
result <- run_full_echogo(
  goseq_file = "input/GOseq_enrichment.tsv",
  trinotate_file = "input/Trinotate.tsv",
  de_file = "input/DE_results.tsv",
  species = c(drerio = "zebrafish", omykiss = "rainbow_trout", hsapiens = "human"),
  target_context = "drerio",
  semantic_reference_orgdb = "org.Dr.eg.db",
  semantic_reference_role = "target_reference",
  run_exploratory_default_domain = FALSE,
  outdir = "results"
)
```

For an organism without a target g:Profiler context, declare
`target_context = NA_character_`. Every requested context then remains an
alternative annotation context while target GOseq remains the experimental
anchor. An omitted or `NULL` value is accepted only through the v0.1.3
compatibility warning path.

# Context roles and domain semantics

The supplied `species` vector is the scientific context panel. It is neither
expanded to a universal panel nor reclassified by phylogeny. When
`target_context` is supplied, that requested context is marked `TARGET`; all
other requested contexts are marked `ALTERNATIVE`.

The researcher may optionally preserve why each context was selected. This is
interpretive provenance, never a statistical input, recurrence weight, evidence
profile rule, display ranking, RRvGO input, network input, or evaluation input.
For example, the following is a synthetic design table, not a claim about any
real study panel:

| context | role | rationale |
|---|---|---|
| `drerio` | `TARGET` | experimental model |
| `species_A` | `ALTERNATIVE` | ecologically relevant to the study question |
| `species_B` | `ALTERNATIVE` | comparative reference |

In an actual analysis, pass only queried g:Profiler contexts:

```{r eval=FALSE}
context_metadata <- data.frame(
  context_code = c("drerio", "omykiss"),
  context_rationale = c(
    "experimental model",
    "relevant to the wider ecological question"
  ),
  selection_basis = c("experimental relevance", "ecological relevance")
)

result <- run_full_echogo(
  input_dir = "input",
  species = c(drerio = "zebrafish", omykiss = "rainbow_trout"),
  target_context = "drerio",
  context_metadata = context_metadata,
  semantic_reference_orgdb = "org.Dr.eg.db",
  semantic_reference_role = "target_reference",
  outdir = "results"
)
```

A model-organism experiment can therefore be explored through contexts chosen
for a wider biological or ecological question. For instance, a zebrafish
fuel-exposure experiment can be interpreted through selected fish references to
ask what existing functions have support or recovery under those contexts and
what new hypotheses deserve follow-up. It does not claim that another fish was
exposed or would respond identically.

Every ordinary v0.1.4 g:Profiler call uses the experiment-derived portable
background and records `background_mode = "custom_experimental_background"`
and `domain_scope = "custom"`. The run metadata records submitted foreground
and background sizes as well as context-effective query/domain sizes.

The optional broader tier is enabled only explicitly:

```{r eval=FALSE}
result <- run_full_echogo(
  goseq_file = "input/GOseq_enrichment.tsv",
  trinotate_file = "input/Trinotate.tsv",
  de_file = "input/DE_results.tsv",
  species = c(hsapiens = "human", mmusculus = "mouse"),
  target_context = NA_character_,
  run_exploratory_default_domain = TRUE,
  run_rrvgo = FALSE,
  outdir = "results_with_default_domain"
)
```

Those calls are labelled `default_domain_exploratory` and use
`domain_scope = "annotated"`. They are saved as separate source provenance and
cannot alter primary exact-term evidence.

# Canonical outputs

The key files are:

| File | Purpose |
|---|---|
| `gprofiler/submitted_vectors/mapping_table.csv` | Original ID, portable name, resolution route, tested/significant membership, and available annotation provenance. |
| `gprofiler/run_manifest.json` | Selected contexts, roles and optional researcher-supplied rationale provenance, plus vector contract, domain modes, submitted/effective sizes, and per-run metadata. |
| `consensus/term_evidence_exact.csv` | Complete exact-term evidence landscape and evidence profiles. |
| `consensus/term_source_provenance_long.csv` | Per-source, per-context adjusted p-values, fold enrichment, contributing genes, recognition/domain metadata, and optional context-rationale provenance. |
| `consensus/context_configuration.csv` | Researcher-selected contexts, their declared roles, and optional supplied rationale provenance. |
| `consensus/annotation_provenance_gene_level.csv` | Available resolver/annotation provenance when an annotation table was supplied. |

The `consensus/` directory and historical
`consensus_enrichment_results_with_and_without_bg.xlsx` filename are retained
for compatibility. They do not indicate consensus scoring: the v0.1.4 content
is exact-term evidence with source-local statistics and provenance.

# Interpret results

`TARGET_ONLY`, `TARGET_PLUS_CONTEXT`, and `ALTERNATIVE_CONTEXT` are the primary
evidence profiles. They are based on source-specific qualification thresholds,
not on row presence or a cross-context p-value calculation.

`ALTERNATIVE_CONTEXT` is a core discovery result: it identifies a function not
qualifying through the target GOseq route but recovered through one or more
alternative contexts using the same custom experimental background. Inspect its
source rows, supplied context-selection rationale, contributing genes,
recognition/domain diagnostics, identifier resolution and annotation provenance
before deciding whether it merits follow-up.

EchoGO supplies a transparent deterministic `display_order` and
`representative_order` only where a display or RRvGO needs an order. Neither is
a confidence score or scientific ranking statistic.

# Compatibility migration

`strict_only` is deprecated. Use `run_exploratory_default_domain = FALSE` for
the normal primary workflow or set it to `TRUE` to request the separate optional
default-domain layer. The fields `consensus_score`, `consensus_score_all`, and
`significant_in_any`, together with legacy `nobg` fields, belong to historical
score-era data and are not canonical v0.1.4 outputs. See
`vignette("migration-v0.1.4", package = "EchoGO")` for the complete public
migration map.
