---
title: "EchoGO Reference-based Workflow: from genome annotation to inputs"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{EchoGO Reference-based Workflow: from genome annotation to inputs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

# What this guide does

This guide starts after alignment, feature counting, and differential-expression
analysis. It turns a gene-level reference RNA-seq project into the input bundle
that EchoGO expects.

You need:

1. a gene-level count matrix,
2. one or more DESeq2 result CSV files,
3. the matching GFF3 or GTF annotation,
4. the matching protein FASTA, and
5. eggNOG-mapper for one external annotation step.

EchoGO does not run the aligner, featureCounts, DESeq2, or eggNOG-mapper.
The exported `echogo_prepare_reference_inputs()` helper validates identifiers,
builds unioned exon-length bias data, creates the expressed-background protein
FASTA, stages eggNOG-mapper, and runs GOseq.

# Install preparation dependencies

Install EchoGO with its vignettes, then install the reference-preparation
packages:

```{r install, eval=FALSE}
remotes::install_github("miloes114/EchoGo", build_vignettes = TRUE)

BiocManager::install(c(
  "Biostrings",
  "rtracklayer",
  "GenomicRanges",
  "IRanges",
  "S4Vectors",
  "goseq"
))
```

HTML reports and vignettes require Pandoc. RStudio includes it. If you run R
from another terminal, check `rmarkdown::pandoc_available()` and install Quarto
or Pandoc if the result is `FALSE`.

If the organism has an OrgDb package, install it too. For zebrafish:

```{r install-orgdb, eval=FALSE}
EchoGO::echogo_install_orgdb("org.Dr.eg.db")
```

# Arrange the source directory

A minimal project can look like this:

```text
reference_project/
|-- allcounts_table.txt
|-- dge_Treated_vs_Control.csv
|-- genome.gff3.gz
`-- proteins.faa.gz
```

The first count-matrix column and the DE table gene column must use the same
identifier type. The annotation and protein FASTA must come from the same
reference release.

A DE table may contain a logical `significant` or `is_de` column. EchoGO
uses that column when present. Otherwise it applies
`padj <= padj_threshold` and
`abs(log2FoldChange) >= log2fc_threshold`.

# Stage 1: build reference maps and the eggNOG FASTA

Use forward slashes in R paths on Windows. The same call is used for both
stages, so keep it in a small function:

```{r prepare-stage-1, eval=FALSE}
library(EchoGO)

prepare_inputs <- function() {
  echogo_prepare_reference_inputs(
    root = "path/to/reference_project",
    gff_file = "genome.gff3.gz",       # GTF/GTF.GZ also works
    protein_fasta = "proteins.faa.gz",
    counts_file = "allcounts_table.txt",
    dge_pattern = "^dge_.*\\.csv$",
    reference_label = "MySpecies",
    orgdb = NULL                       # see the OrgDb section below
  )
}

prep <- prepare_inputs()
prep$status
# "awaiting_emapper"

readLines(prep$emapper_command)
```

Stage 1 writes reusable files under `echogo_work/`:

- `02_maps/gene_protein_map.tsv`
- `02_maps/gene_lengths.tsv`
- `03_proteins/BG_universe_proteins.fa`
- `04_emapper/RUN_EMAPPER.txt`

Gene-length bias uses the union of exon intervals per gene. If exon records are
absent, CDS intervals are used and this fallback is reported.

Large vertebrate annotations can require several gigabytes of memory and a few
minutes on the first import. The next call reuses the reference maps and FASTA.
Set `reuse_reference = FALSE` only after changing the annotation, FASTA, or ID
rules.

# External step: run eggNOG-mapper

Run the command saved in `prep$emapper_command` in an environment where
`emapper.py` is installed. A typical command is:

```bash
emapper.py \
  -i "echogo_work/03_proteins/BG_universe_proteins.fa" \
  --itype proteins \
  -m diamond \
  --output BG_universe \
  --output_dir "echogo_work/04_emapper" \
  --cpu 8
```

Local installations may also require `--data_dir`. The required result is:

```text
echogo_work/04_emapper/BG_universe.emapper.annotations
```

If you already have a compatible annotation file, pass its absolute path as
`emapper_file`; EchoGO will not rerun eggNOG-mapper.

# Stage 2: create the EchoGO bundle and run GOseq

After the eggNOG annotation exists, rerun the same call:

```{r prepare-stage-2, eval=FALSE}
prep <- prepare_inputs()
stopifnot(identical(prep$status, "complete"))

prep$output_dir
prep$summary
```

The completed `echogo_input/` directory contains one set per detected
contrast:

| File | Purpose |
|---|---|
| `allcounts_table.txt` | Gene-level count background |
| `dge_<CONTRAST>.csv` | DESeq2 result table |
| `dge_<CONTRAST>.GOseq.enriched.tsv` | GOseq terms passing adjusted FDR 0.05 |
| `Trinotate_for_EchoGO.tsv` | Minimal gene-level annotation bridge |
| `<reference_label>_eggNOG_for_EchoGO.tsv` | Labelled eggNOG annotation |

The generated annotation keeps semantic fields separate. An eggNOG preferred
name or seed ortholog is never copied into `sprot_Top_BLASTX_hit`. EchoGO
submits a genuine SwissProt gene symbol first when present, otherwise an
eggNOG preferred name, otherwise a portable native symbol. Raw reference,
transcript, contig, seed-ortholog, and accession IDs remain visible in the
mapping provenance but are not sent to g:Profiler. Taxonomy labels are retained
for provenance and are not interpreted as an ordered numeric cutoff.

The GOseq file is validated to contain exactly the seven columns EchoGO needs:
`category`, `term`, `ontology`, `numDEInCat`, `numInCat`,
`over_represented_FDR`, and `gene_ids`.

# Choose the GO source

For a model organism with a well-supported OrgDb, use the species-native
mapping for GOseq and retain eggNOG for protein names and organism-context
annotation. This mirrors the validated GRCz11 zebrafish workflow:

```{r native-orgdb, eval=FALSE}
prep <- echogo_prepare_reference_inputs(
  root = "path/to/reference_project",
  gff_file = "genome.gtf.gz",
  protein_fasta = "proteins.faa.gz",
  emapper_file = "path/to/annotations/reference.emapper.annotations",
  reference_label = "drerio",
  orgdb = "org.Dr.eg.db"
)
```

For a non-model organism without an OrgDb, leave `orgdb = NULL`. GOseq then
uses GO terms recovered from eggNOG-mapper.

Do not substitute an OrgDb from another organism merely to make the function
run. Its gene identifiers will not represent the native background.

# Run EchoGO on one contrast

Explicit filenames are safest when an input directory contains more than one
contrast:

```{r run-echogo, eval=FALSE}
input_dir <- "path/to/reference_project/echogo_input"

result <- run_full_echogo(
  input_dir = input_dir,
  goseq_file = file.path(
    input_dir,
    "dge_Treated_vs_Control.GOseq.enriched.tsv"
  ),
  trinotate_file = file.path(input_dir, "Trinotate_for_EchoGO.tsv"),
  de_file = file.path(input_dir, "dge_Treated_vs_Control.csv"),
  count_matrix_file = file.path(input_dir, "allcounts_table.txt"),
  species = c("hsapiens", "mmusculus", "drerio"),
  target_context = "drerio",
  semantic_reference_orgdb = "org.Dr.eg.db",
  semantic_reference_role = "target_reference",
  outdir = "path/to/reference_project/echogo_results",
  run_rrvgo = TRUE,
  make_report = TRUE,
  verbose = TRUE
)
```

The `semantic_reference_orgdb` passed to `run_full_echogo()` controls RRvGO
semantic similarity only. It is conceptually separate from the optional
species-native `orgdb` used during input preparation, even when the same target
OrgDb package is appropriate for both purposes. `orgdb` remains a deprecated
v0.1.4 compatibility alias for the RRvGO argument.

# Troubleshooting

## No DE genes overlap the count background

Check the identifier examples before changing thresholds:

```{r inspect-ids, eval=FALSE}
counts <- read.delim("allcounts_table.txt", check.names = FALSE)
dge <- read.csv("dge_Treated_vs_Control.csv", check.names = FALSE)

head(counts[[1]])
head(dge[[1]])
```

Common causes are version suffixes, `gene-` prefixes, symbols mixed with
accessions, or DE files from a different reference release.

## Few genes map to proteins or lengths

The helper stops below 50% coverage. Verify that the GFF3/GTF and protein FASTA
belong together. For unusual annotations, set `counts_id_column` or
`dge_id_column` explicitly. Set `strip_gene_prefix = FALSE` when
`gene-` is part of the real count identifier.

## Protein IDs do not match eggNOG queries

The `#query` values in the eggNOG file must be the FASTA identifiers used in
stage 1. Do not annotate a different proteome or rename FASTA headers between
stages.

## A rerun is importing the full reference again

Keep the same `work_dir` and leave `reuse_reference = TRUE`. The cache is
invalidated manually with `reuse_reference = FALSE` after a reference change.

## No GOseq terms pass FDR

This can be a valid result. Check `prep$summary`, the number of DE genes after
length/GO filtering, and whether the selected OrgDb matches the organism.
Do not switch from adjusted FDR to raw p-values to force results.

# Next steps

See `vignette("EchoGO_workflow")` for the complete pipeline and
`vignette("EchoGO_interpretation")` for exact-term evidence, RRvGO, and network
interpretation. Use `citation("EchoGO")` for the package citation.
