---
title: "EchoGO biological interpretation report"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: false
    self_contained: true
    code_folding: hide
    highlight: tango
    css: echogo_report.css
params:
  dirs: NULL
  top_n: 25
  key_findings_max_target_recovered: 8
  key_findings_max_target_only: 4
  key_findings_max_hypotheses: 8
  landscape_max_terms_per_profile: 10
  drilldown_term_id: null
  sections: ["overview", "representability", "goseq", "gprofiler", "landscape", "rrvgo", "networks", "drilldown", "exploration", "technical"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center")
suppressPackageStartupMessages({
  library(EchoGO)
  library(readr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(tibble)
  library(stringr)
})

`%||%` <- function(x, y) if (is.null(x)) y else x
safe_n <- function(x) if (is.null(x)) 0L else length(x)
fmt_n <- function(x) format(as.numeric(x), big.mark = ",", scientific = FALSE, trim = TRUE)
first_or <- function(x, fallback = NA_character_) {
  x <- x[!is.na(x) & nzchar(as.character(x))]
  if (length(x)) as.character(x[[1]]) else fallback
}
read_table <- function(path) {
  if (is.null(path) || length(path) == 0L || is.na(path) || !file.exists(path)) return(tibble::tibble())
  if (grepl("\\.xlsx$", path, ignore.case = TRUE)) return(tibble::as_tibble(openxlsx::read.xlsx(path)))
  suppressMessages(readr::read_csv(path, show_col_types = FALSE))
}
find_first <- function(paths) {
  hit <- paths[file.exists(paths)]
  if (length(hit)) normalizePath(hit[[1]], winslash = "/", mustWork = FALSE) else NA_character_
}
find_recursive <- function(root, pattern) {
  if (!dir.exists(root)) return(character())
  list.files(root, pattern = pattern, recursive = TRUE, full.names = TRUE, ignore.case = TRUE)
}

metric_cards <- getFromNamespace(".echogo_report_metric_cards", "EchoGO")
callout <- getFromNamespace(".echogo_report_callout", "EchoGO")
figure_card <- getFromNamespace(".echogo_report_figure_card", "EchoGO")
table_details <- getFromNamespace(".echogo_report_table_details", "EchoGO")
stage_asset <- getFromNamespace(".echogo_stage_asset", "EchoGO")
semantic_badge <- getFromNamespace(".echogo_report_semantic_badge", "EchoGO")
save_variants <- getFromNamespace(".echogo_save_plot_variants", "EchoGO")
plot_representability <- getFromNamespace("plot_echogo_representability", "EchoGO")
plot_profile_counts <- getFromNamespace("plot_echogo_profile_counts", "EchoGO")
plot_context_summary <- getFromNamespace("plot_echogo_context_summary", "EchoGO")
plot_landscape <- getFromNamespace("plot_echogo_evidence_landscape", "EchoGO")
plot_trace <- getFromNamespace("plot_echogo_hypothesis_trace", "EchoGO")
plot_network <- getFromNamespace("plot_echogo_gene_overlap_network", "EchoGO")
mapping_summary_fun <- getFromNamespace(".echogo_mapping_summary", "EchoGO")
key_findings_data_fun <- getFromNamespace(".echogo_key_findings_data", "EchoGO")
plot_key_findings <- getFromNamespace("plot_echogo_key_findings_map", "EchoGO")
primary_metric_cards <- getFromNamespace(".echogo_primary_metric_cards", "EchoGO")
representability_statement <- getFromNamespace(".echogo_representability_statement", "EchoGO")
report_logo_html <- getFromNamespace(".echogo_report_logo_html", "EchoGO")
discover_rrvgo <- getFromNamespace(".echogo_discover_rrvgo_products", "EchoGO")
read_rrvgo_status <- getFromNamespace(".echogo_read_rrvgo_status", "EchoGO")
rrvgo_status_message <- getFromNamespace(".echogo_rrvgo_status_message", "EchoGO")
read_network_status <- getFromNamespace(".echogo_read_network_status", "EchoGO")
gprofiler_report_status <- getFromNamespace(".echogo_gprofiler_report_status", "EchoGO")

dirs <- params$dirs %||% list()
base_dir <- normalizePath(dirs$base %||% getwd(), winslash = "/", mustWork = FALSE)
report_dir <- normalizePath(dirs$report %||% file.path(base_dir, "report"), winslash = "/", mustWork = FALSE)
dir.create(report_dir, recursive = TRUE, showWarnings = FALSE)
generated_dir <- file.path(report_dir, "generated")
generated_assets_dir <- file.path(report_dir, "assets", "generated")
dir.create(generated_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(generated_assets_dir, recursive = TRUE, showWarnings = FALSE)

consensus_dir <- file.path(base_dir, "consensus")
gprof_dir <- file.path(base_dir, "gprofiler")
goseq_dir <- file.path(base_dir, "goseq")
rrvgo_dir <- file.path(base_dir, "rrvgo")
network_dir <- file.path(base_dir, "networks")
diagnostics_dir <- file.path(base_dir, "diagnostics")

evidence_path <- find_first(c(
  file.path(consensus_dir, "term_evidence_exact.csv"),
  file.path(consensus_dir, "term_evidence_exact.xlsx")
))
sources_path <- file.path(consensus_dir, "term_source_provenance_long.csv")
contexts_path <- file.path(consensus_dir, "context_configuration.csv")
mapping_path <- file.path(gprof_dir, "submitted_vectors", "mapping_table.csv")
annotation_path <- file.path(consensus_dir, "annotation_provenance_gene_level.csv")
config_path <- file.path(diagnostics_dir, "run_configuration.json")

evidence <- read_table(evidence_path)
sources <- read_table(sources_path)
contexts <- read_table(contexts_path)
mapping <- read_table(mapping_path)
annotation <- read_table(annotation_path)
configuration <- tryCatch(jsonlite::read_json(config_path, simplifyVector = TRUE), error = function(e) list())

if (nrow(evidence)) {
  evidence <- evidence |>
    mutate(
      ontology = case_when(
        .data$ontology %in% c("GO:BP", "BP") ~ "BP",
        .data$ontology %in% c("GO:MF", "MF") ~ "MF",
        .data$ontology %in% c("GO:CC", "CC") ~ "CC",
        TRUE ~ as.character(.data$ontology)
      )
    )
}
if (nrow(sources) && "ontology" %in% names(sources)) {
  sources <- sources |>
    mutate(
      ontology = case_when(
        .data$ontology %in% c("GO:BP", "BP") ~ "BP",
        .data$ontology %in% c("GO:MF", "MF") ~ "MF",
        .data$ontology %in% c("GO:CC", "CC") ~ "CC",
        TRUE ~ as.character(.data$ontology)
      )
    )
}

map_summary <- mapping_summary_fun(mapping)
n_target_only <- if (nrow(evidence)) sum(evidence$evidence_profile == "TARGET_ONLY", na.rm = TRUE) else 0L
n_target_plus <- if (nrow(evidence)) sum(evidence$evidence_profile == "TARGET_PLUS_CONTEXT", na.rm = TRUE) else 0L
n_alt <- if (nrow(evidence)) sum(evidence$evidence_profile == "ALTERNATIVE_CONTEXT", na.rm = TRUE) else 0L
n_target_supported <- n_target_only + n_target_plus
n_contexts <- nrow(contexts)
n_alt_contexts <- if (nrow(contexts)) sum(contexts$context_role == "ALTERNATIVE", na.rm = TRUE) else 0L
target_context_label <- if (nrow(contexts) && any(contexts$context_role == "TARGET")) {
  first_or(contexts$context_label[contexts$context_role == "TARGET"], "not declared")
} else {
  "none declared"
}

hero_cards <- primary_metric_cards(evidence, contexts)
logo_src <- if (file.exists(file.path(base_dir, "echogo-logo.png"))) "echogo-logo.png" else NA_character_

# Generate universal summary plots from the frozen evidence/provenance tables.
plot_assets <- list()
safe_save <- function(plot, stem, width = 12, height = 8) {
  if (is.null(plot)) return(list(png = NA_character_, pdf = NA_character_, svg = NA_character_))
  tryCatch(
    save_variants(plot, generated_dir, stem, width = width, height = height),
    error = function(e) list(png = NA_character_, pdf = NA_character_, svg = NA_character_)
  )
}
plot_assets$profiles <- safe_save(plot_profile_counts(evidence), "evidence_profile_summary", 10, 6)
plot_assets$representability <- safe_save(plot_representability(mapping), "representability_flow", 12, 6)
plot_assets$contexts <- safe_save(plot_context_summary(sources, evidence, contexts), "reference_species_summary", 15, max(6, 0.48 * max(4, n_contexts)))
key_findings_tables <- list()
for (ont in c("BP", "MF", "CC")) {
  key_findings_tables[[ont]] <- tryCatch(
    key_findings_data_fun(
      evidence,
      ontology = ont,
      max_target_recovered = params$key_findings_max_target_recovered %||% 8L,
      max_target_only = params$key_findings_max_target_only %||% 4L,
      max_hypotheses = params$key_findings_max_hypotheses %||% 8L
    ),
    error = function(e) tibble::tibble()
  )
  if (nrow(key_findings_tables[[ont]])) {
    readr::write_csv(key_findings_tables[[ont]], file.path(generated_assets_dir, paste0("key_findings_", ont, ".csv")))
  }
  kp <- tryCatch(
    plot_key_findings(
      evidence,
      ontology = ont,
      max_target_recovered = params$key_findings_max_target_recovered %||% 8L,
      max_target_only = params$key_findings_max_target_only %||% 4L,
      max_hypotheses = params$key_findings_max_hypotheses %||% 8L
    ),
    error = function(e) NULL
  )
  plot_assets[[paste0("key_findings_", ont)]] <- safe_save(kp, paste0("key_findings_", ont), 14, 10)
}
for (ont in c("BP", "MF", "CC")) {
  p <- tryCatch(
    plot_landscape(
      evidence, sources, contexts, ont,
      max_terms_per_profile = params$landscape_max_terms_per_profile %||% 10L
    ),
    error = function(e) NULL
  )
  plot_assets[[paste0("landscape_", ont)]] <- safe_save(
    p, paste0("evidence_landscape_", ont),
    width = 16,
    height = max(7, 0.34 * sum(evidence$ontology == ont & evidence$evidence_profile %in% c("TARGET_ONLY", "TARGET_PLUS_CONTEXT", "ALTERNATIVE_CONTEXT"), na.rm = TRUE) + 4)
  )
}

# Plain biological name summaries; deterministic and intentionally non-inferential.
example_terms <- function(profile, n = 3L) {
  if (!nrow(evidence)) return(character())
  x <- evidence |>
    filter(.data$evidence_profile %in% profile, !is.na(.data$term_name), nzchar(.data$term_name)) |>
    arrange(.data$display_order, .data$term_id) |>
    distinct(.data$term_name) |>
    slice_head(n = n) |>
    pull(.data$term_name)
  x
}
format_examples <- function(x) {
  if (!length(x)) return("No representative term names were available for this summary.")
  paste(x, collapse = "; ")
}

target_examples <- example_terms(c("TARGET_ONLY", "TARGET_PLUS_CONTEXT"))
hypothesis_examples <- example_terms("ALTERNATIVE_CONTEXT")

# Source plot lookup. The pipeline writes current plots to canonical directories;
# compatibility directories are accepted only as a fallback.
plot_sidecars <- function(path) {
  if (is.na(path) || !nzchar(path)) return(list(png = NA_character_, pdf = NA_character_, svg = NA_character_))
  stem <- tools::file_path_sans_ext(path)
  list(
    png = if (file.exists(paste0(stem, ".png"))) paste0(stem, ".png") else NA_character_,
    pdf = if (file.exists(paste0(stem, ".pdf"))) paste0(stem, ".pdf") else NA_character_,
    svg = if (file.exists(paste0(stem, ".svg"))) paste0(stem, ".svg") else NA_character_
  )
}
find_goseq_plot <- function(ont) {
  hits <- find_recursive(goseq_dir, paste0("GO_lollipop_allDepths_top50_", ont, "\\.(png|pdf)$"))
  if (!length(hits)) return(NA_character_)
  png <- hits[grepl("\\.png$", hits, ignore.case = TRUE)]
  if (length(png)) png[[1]] else hits[[1]]
}
find_gp_plot <- function(label, ont, exploratory = FALSE) {
  roots <- if (exploratory) {
    c(file.path(gprof_dir, "default_domain_exploratory"), file.path(gprof_dir, "no_background_genome_wide"))
  } else {
    c(file.path(gprof_dir, "custom_experimental_background"), file.path(gprof_dir, "with_custom_background"))
  }
  escaped <- gsub("([][{}()+*^$|\\?.])", "\\\\\\1", label)
  patterns <- c(
    paste0("gprofiler_", escaped, ".*", ont, ".*\\.(png|pdf)$"),
    paste0(escaped, ".*", ont, ".*lollipop.*\\.(png|pdf)$")
  )
  for (root in roots) {
    for (pat in patterns) {
      hits <- find_recursive(root, pat)
      if (length(hits)) {
        png <- hits[grepl("\\.png$", hits, ignore.case = TRUE)]
        return(if (length(png)) png[[1]] else hits[[1]])
      }
    }
  }
  NA_character_
}

rrvgo_discovery <- discover_rrvgo(rrvgo_dir, evidence, configuration)
rrvgo_cluster_files <- rrvgo_discovery$cluster_files
rrvgo_clusters <- if (length(rrvgo_cluster_files)) {
  bind_rows(lapply(rrvgo_cluster_files, function(path) {
    x <- read_table(path)
    x$cluster_file <- path
    x
  }))
} else tibble::tibble()

rrvgo_status <- read_rrvgo_status(rrvgo_dir)
network_status <- read_network_status(network_dir)
semantic_orgdb <- rrvgo_discovery$semantic_reference_orgdb %||% configuration$semantic_reference_orgdb %||% NA_character_
semantic_role <- rrvgo_discovery$semantic_reference_role %||% configuration$semantic_reference_role %||% NA_character_
semantic_method <- rrvgo_discovery$semantic_method %||% configuration$semantic_method %||% "Rel"
run_rrvgo <- isTRUE(rrvgo_discovery$enabled)
run_exploratory <- isTRUE(configuration$run_exploratory_default_domain)
rrvgo_status_for <- function(product, ont) {
  if (!is.data.frame(rrvgo_status) || !nrow(rrvgo_status)) return(tibble::tibble())
  dplyr::filter(rrvgo_status, .data$semantic_product == product, toupper(.data$ontology) == toupper(ont))
}
```

<div class="hero-echogo">
<div class="hero-brand-row">
<div class="hero-copy">
<div class="kicker">EchoGO v0.1.4</div>
<h1>Traceable functional interpretation across annotation contexts</h1>
<div class="subtitle">Biological interpretation report</div>
<div class="hero-rule"></div>
<p>Start with your target analysis, see which findings are also recovered by selected annotation contexts, and inspect the context-derived hypotheses.</p>
<div class="hero-meta">
<span><strong>Target context in g:Profiler:</strong> `r target_context_label`</span>
<span><strong>Chosen annotation contexts:</strong> `r n_contexts`</span>
<span><strong>Alternative annotation contexts:</strong> `r n_alt_contexts`</span>
</div>
</div>
`r report_logo_html(logo_src)`
</div>
</div>

```{r hero-cards, results='asis'}
cat(metric_cards(hero_cards))
```

<div class="report-scope"><strong>What EchoGO is doing</strong>EchoGO interprets one transcriptomic experiment using selected annotation contexts. GOseq from the target analysis remains the experimental anchor. These organism-specific annotation resources are not additional experiments; recovery of the same term does not show that the corresponding organisms share a biological response.</div>
<!-- Generic compatibility anchors retained for downstream contract checks:
Experiment and representability; What the target analysis already says;
New alternative-context hypotheses; Target-supported semantic themes;
Alternative-context hypothesis semantic themes; Hypothesis drill-down;
Default-domain exploratory hypotheses. The visible report uses the universal
language headings defined by the current amendment. -->

# 1. Your results at a glance

Start here. EchoGO separates three biological questions that are easy to mix when reading enrichment tables: **what your target analysis supports, which target findings are also recovered by selected alternative annotation contexts, and which context-derived hypotheses merit follow-up.**

```{r glance-how, results='asis'}
cat(callout(
  "how-to-read", "How to read this section",
  "The counts describe the structure of your results, not biological quality. More recovered terms or more context-derived hypotheses do not make one annotation context better than another."
))
cat(figure_card(
  "Primary EchoGO result structure",
  plot_assets$profiles$png,
  "The three bars separate terms supported by the target analysis, target findings also recovered in selected annotation contexts, and context-derived hypotheses.",
  report_dir, base_dir,
  plot_assets$profiles$pdf, plot_assets$profiles$svg
))
```

```{r glance-text, results='asis'}
cat(callout(
  "takeaway", "What we see",
  paste0(
    "Your primary result contains <strong>", fmt_n(n_target_only),
    " target-only terms</strong>, <strong>", fmt_n(n_target_plus),
    " target findings also recovered in selected annotation contexts</strong>, and <strong>",
    fmt_n(n_alt), " context-derived hypotheses</strong>."
  )
))
cat(callout(
  "meaning", "Biological reading",
  paste0(
    "The target-supported terms are the biological story already anchored in your experiment. Examples among the first displayed target-supported terms include: <em>",
    format_examples(target_examples),
    "</em>. Context-derived hypothesis evidence is kept separate; examples include: <em>",
    format_examples(hypothesis_examples), "</em>."
  )
))
cat(callout(
  "caution", "Keep in mind",
  "EchoGO does not create an integrated significance score. Adjusted p-values and fold enrichment remain attached to the GOseq or g:Profiler analysis that produced them."
))
```

# 2. Key functional findings

This is the compact EchoGO results view. It separates functions already supported by your target analysis from context-derived hypotheses that appear when the same experiment is interpreted through selected annotation contexts. The position of each term shows how many selected alternative contexts recover that exact GO term. This is descriptive context recurrence, not a new significance score.

```{r key-findings-how, results='asis'}
cat(callout(
  "how-to-read", "How to read these figures",
  "Each row is one exact GO term. Terms supported by your target analysis remain target findings whether or not a selected annotation context also reports them. For terms recovered in alternative contexts, the n/N label shows how many selected contexts recover the same exact term using the matched experimental background. Point size shows the number of unique contributing genes reported across the retained source evidence. The figure is a compact display; the complete evidence table and every context-specific result remain available later in the report."
))
cat(callout(
  "takeaway", "What we see",
  paste0(
    "Your target analysis supports <strong>", fmt_n(n_target_supported),
    " exact GO terms</strong>. Of these, <strong>", fmt_n(n_target_plus),
    " are also recovered through at least one of the selected alternative annotation contexts</strong>, while <strong>",
    fmt_n(n_target_only), " remain target-only within this panel</strong>. The selected alternative annotation contexts additionally produce <strong>",
    fmt_n(n_alt), " exact GO terms retained as new functional hypotheses</strong>."
  )
))
cat(callout(
  "meaning", "Interpretation",
  "Target findings define the biological story already anchored in your experiment. Recovery through selected alternative annotation contexts shows which parts of that story are also found through those resources. Context-derived terms broaden the set of ideas worth investigating, but remain a separate hypothesis layer until supported by evidence from the target experiment or other biological data."
))
cat(callout(
  "caution", "Keep in mind",
  "A term recovered by many annotation contexts is not automatically more significant, more conserved or more likely to be true. The count only describes how broadly the same exact term appears across the selected context panel."
))
for (ont in c("BP", "MF", "CC")) {
  a <- plot_assets[[paste0("key_findings_", ont)]]
  cat(figure_card(
    paste0("Key Findings Map - ", ont),
    a$png,
    "The map is a compact display of exact-term evidence. Point position and the n/N label show literal recovery across the selected alternative annotation contexts; point size is a descriptive contributing-gene count.",
    report_dir, base_dir,
    a$pdf, a$svg
  ))
}
cat(EchoGO:::.echogo_html_details(
  "How were terms chosen for this compact display?",
  "Target findings also recovered by alternative contexts are displayed first by the number of selected contexts recovering the exact term; ties use the target GOseq adjusted p-value only within the target analysis. Target-only findings use the target GOseq adjusted p-value. Context-derived hypotheses are displayed by context recurrence and deterministic term order; adjusted p-values from different annotation contexts are never combined or compared to create a cross-context ranking. All primary terms remain available in the complete Evidence Landscape and evidence table."
))
```

# 3. How much of your experiment could be compared across annotation contexts?

```{r representability-statement, results='asis'}
statement <- representability_statement(mapping)
if (!is.na(statement) && nzchar(statement)) cat(callout("takeaway", "Representability", statement))
```

Before comparing annotation contexts, EchoGO checks how much of the tested target-gene universe can be represented by the portable names submitted to g:Profiler. Foreground and background use the same resolver, and the resulting vectors are reused across selected contexts.

```{r representability-figure, results='asis'}
cat(callout(
  "how-to-read", "How to read this figure",
  "The upper lane follows the complete tested target universe into the unique names available for comparison. The lower lane follows significant target entities into the submitted foreground. Identifier loss here is different from context-specific recognition later in the report."
))
cat(figure_card(
  "From your experiment to the annotation-context comparison",
  plot_assets$representability$png,
  "This is a representability audit. Missing names or duplicate-name collapse reduce what can enter the comparison, but they do not mean that excluded target genes lack biological function.",
  report_dir, base_dir,
  plot_assets$representability$pdf, plot_assets$representability$svg
))
```

```{r representability-text, results='asis'}
if (nrow(map_summary)) {
  cat(callout(
    "takeaway", "What we see",
    paste0(
      "Of <strong>", fmt_n(map_summary$tested_entities), " tested target entities</strong>, <strong>",
      fmt_n(map_summary$submitted_background_names), " unique names</strong> entered the matched-background comparison. The mapping audit records <strong>",
      fmt_n(map_summary$tested_without_portable_name), " tested entities without a portable name</strong> and <strong>",
      fmt_n(map_summary$tested_duplicate_rows_collapsed), " duplicate-name rows collapsed</strong>. The significant foreground moved from <strong>",
      fmt_n(map_summary$significant_entities), " target entities</strong> to <strong>",
      fmt_n(map_summary$submitted_foreground_names), " unique submitted names</strong>."
    )
  ))
}
cat(callout(
  "meaning", "What this means",
  "This step defines how much of your experiment is available for comparison across annotation contexts. When this fraction is small, identifier representability limits what each selected context can reveal."
))
cat(callout(
  "caution", "Keep in mind",
  "Identifier loss and context-specific recognition are different processes. This figure shows what could be submitted; the next summary shows how much of that submitted query each selected context recognized."
))
```

```{r representability-details, results='asis'}
mapping_url <- if (file.exists(mapping_path)) stage_asset(mapping_path, report_dir, base_dir) else NULL
cat(table_details("Mapping and representability audit", mapping, mapping_url, preview_rows = params$top_n))
if (nrow(annotation)) cat(table_details("Gene-level annotation provenance", annotation, NULL, preview_rows = params$top_n))
if (nrow(contexts)) cat(table_details("Selected annotation contexts and declared roles", contexts, NULL, preview_rows = params$top_n))
```

# 4. What does your target analysis support?

GOseq is the target-analysis anchor in EchoGO. It tests enrichment using your target experiment and its tested-gene universe while accounting for RNA-seq gene-length bias. Read this section first as the biological interpretation available **before alternative annotation contexts add information**.

```{r goseq-how, results='asis'}
cat(callout(
  "how-to-read", "How to read these figures",
  "Each figure contains GOseq results from your target analysis only. Fold enrichment describes over-representation in GOseq, colour represents the source-local GOseq adjusted significance used by the established plot, and point size represents contributing target genes where available. These statistics are not combined with g:Profiler statistics later."
))
for (ont in c("BP", "MF", "CC")) {
  p <- find_goseq_plot(ont)
  side <- plot_sidecars(p)
  if (!is.na(p) && grepl("\\.png$", p, ignore.case = TRUE)) side$png <- p
  cat(figure_card(
    paste0("Target GOseq enrichment - ", ont),
    side$png,
    paste0(
      "Source-local GOseq enrichment for ", ont,
      ". The exact qualifying terms are carried forward unchanged into the EchoGO evidence table."
    ),
    report_dir, base_dir,
    side$pdf, side$svg
  ))
}
```

```{r goseq-reading, results='asis'}
cat(callout(
  "takeaway", "What we see",
  paste0(
    "Your target analysis supports <strong>", fmt_n(n_target_supported),
    " exact GO terms</strong> in the primary EchoGO result. Examples among the first displayed target-supported terms include: <em>",
    format_examples(target_examples), "</em>."
  )
))
cat(callout(
  "meaning", "What this means",
  "These terms answer the first EchoGO question: what does the target analysis support? The next sections ask whether the same exact findings are also recovered through selected annotation contexts and which context-derived hypotheses emerge beyond the target result."
))
cat(callout(
  "caution", "Keep in mind",
  "A target GOseq term can remain biologically important even if no alternative annotation context recovers it. EchoGO keeps that result explicitly as target-only rather than treating it as weaker combined evidence."
))
```

# 5. What do your selected annotation contexts recover?

EchoGO evaluates the same resolved experimental foreground and matched background independently using each selected g:Profiler annotation context. These organism-specific resources provide functional annotations for interpreting your experiment; EchoGO does not construct a different experimental gene list for each context or rank contexts as better or worse.

```{r reference-summary, results='asis'}
if (identical(target_context_label, "not declared")) {
  cat(callout(
  "method-note", "Target context",
    "No target g:Profiler context was declared for this run. Target GOseq remains the experimental anchor, and all queried contexts are treated as selected alternatives."
  ))
} else {
  cat(callout(
    "method-note", "Target context",
    paste0("The run declares <strong>", target_context_label, "</strong> as the target context in g:Profiler. Target GOseq remains the experimental enrichment anchor.")
  ))
}
cat(callout(
  "how-to-read", "How to read the annotation-context summary",
  "The first panel shows how much of the submitted foreground each context recognized. The second shows how many target-supported exact GO terms each context also recovered. The third shows how many context-derived hypothesis terms it added. Contexts remain in the declared order and are not ranked."
))
cat(figure_card(
  "What your selected annotation contexts recover",
  plot_assets$contexts$png,
  "The same experimental query is compared across selected contexts. Differences can reflect identifier recognition, annotation depth and the statistical domain available in each resource.",
  report_dir, base_dir,
  plot_assets$contexts$pdf, plot_assets$contexts$svg
))
cat(callout(
  "meaning", "What this means",
  "An annotation context can recover part of the target-supported evidence and can also add terms that were not qualifying target GOseq findings. Those outcomes are kept separate in the Evidence Landscape below."
))
cat(callout(
  "caution", "Keep in mind",
  "A context that returns more terms is not automatically better, biologically closer or more trustworthy. Identifier recognition and annotation depth influence the number of terms recovered."
))
```

## Source-level g:Profiler plots by annotation context

Open a context below to inspect its g:Profiler enrichment plots. **Each adjusted p-value and fold-enrichment value belongs only to the analysis that produced it.**

```{r gprofiler-gallery, results='asis'}
if (!nrow(contexts)) {
  cat("<em>No g:Profiler annotation contexts were configured for this run.</em>")
} else {
  for (i in seq_len(nrow(contexts))) {
    label <- contexts$context_label[[i]]
    role <- contexts$context_role[[i]]
    badge_class <- if (identical(role, "TARGET")) "target" else "alternative"
    badge_text <- if (identical(role, "TARGET")) "Target context" else "Alternative context"
    inner <- paste0("<div><span class='context-badge ", badge_class, "'>", badge_text, "</span></div>")
    for (ont in c("BP", "MF", "CC")) {
      gp_status <- gprofiler_report_status(sources, label, ont, exploratory = FALSE)
      p <- find_gp_plot(label, ont, exploratory = FALSE)
      if (identical(gp_status$status[[1]], "FIGURE_EXPECTED") && is.na(p)) {
        stop(
          paste0("g:Profiler report asset missing for ", label, " ", ont,
                 " although qualifying source rows were recorded."),
          call. = FALSE
        )
      }
      side <- plot_sidecars(p)
      if (!is.na(p) && grepl("\\.png$", p, ignore.case = TRUE)) side$png <- p
      if (gp_status$status[[1]] %in% c("NO_SOURCE_ROWS", "NO_QUALIFYING_TERMS")) {
        inner <- paste0(
          inner,
          "<div class='empty-state'><strong>", gp_status$status[[1]],
          "</strong><br/>", gp_status$message[[1]], "</div>"
        )
      } else {
        inner <- paste0(
          inner,
          figure_card(
            paste0(label, " - matched-background g:Profiler - ", ont),
            side$png,
            paste0(
              "The same experiment-derived foreground and matched background were evaluated using ",
              label,
              " as the annotation context. The plotted adjusted p-values and fold enrichment belong only to this source analysis."
            ),
            report_dir, base_dir,
            side$pdf, side$svg
          )
        )
      }
    }
    cat(paste0("<details><summary>", label, "</summary><div style='margin-top:12px;'>", inner, "</div></details>"))
  }
}
```

# 6. Where does each finding come from?

## EchoGO Evidence Landscape

The Key Findings Map summarizes what deserves inspection. The Evidence Landscape below shows the complete source pattern behind those findings. Each column is your target GOseq analysis or one selected annotation context, so you can see exactly which analyses recover each displayed term.

```{r landscape-how, results='asis'}
cat(callout(
  "how-to-read", "How to read the Evidence Landscape",
  "Each row is one exact GO term. Target GOseq is the experimental anchor. The remaining columns are selected g:Profiler annotation contexts using the matched experimental background. A filled cell means the term qualified in that source or context. The bar at right shows how many selected alternative contexts recovered that exact term."
))
for (ont in c("BP", "MF", "CC")) {
  a <- plot_assets[[paste0("landscape_", ont)]]
  cat(figure_card(
    paste0("EchoGO Evidence Landscape - ", ont),
    a$png,
    "Filled cells indicate qualifying source-local evidence. The right-hand bar is descriptive recurrence across the selected alternative annotation contexts; it is not a probability or a combined p-value.",
    report_dir, base_dir,
    a$pdf, a$svg
  ))
}
```

<div class="profile-definitions">
<div class="profile-definition"><strong>Target only</strong><code>TARGET_ONLY</code><br/>Supported by your target GOseq analysis and not recovered in selected alternative annotation contexts.</div>
<div class="profile-definition"><strong>Target + context</strong><code>TARGET_PLUS_CONTEXT</code><br/>Supported by your target GOseq analysis and also recovered in at least one selected alternative annotation context.</div>
<div class="profile-definition"><strong>Context-derived hypotheses</strong><code>ALTERNATIVE_CONTEXT</code><br/>Not a qualifying target GOseq term, but recovered in at least one selected alternative annotation context.</div>
</div>

```{r landscape-reading, results='asis'}
cat(callout(
  "meaning", "What this means",
  "The matrix exposes how dependent the interpretation is on the annotation contexts without creating a new significance statistic. Your target-supported results remain anchored in the experiment; terms found only through alternative contexts remain a separate hypothesis layer."
))
cat(callout(
  "caution", "Keep in mind",
  "The number of alternative contexts recovering a term is not evidence of evolutionary conservation, independent replication, orthology validation or probability of truth. It records how often the same experimental query recovered the exact term across the selected contexts."
))
```

```{r evidence-details, results='asis'}
evidence_url <- if (!is.na(evidence_path) && file.exists(evidence_path)) stage_asset(evidence_path, report_dir, base_dir) else NULL
sources_url <- if (file.exists(sources_path)) stage_asset(sources_path, report_dir, base_dir) else NULL
cat(table_details("Complete EchoGO exact-term evidence", evidence, evidence_url, preview_rows = params$top_n))
cat(table_details("Complete source-by-context provenance", sources, sources_url, preview_rows = params$top_n))
```

# 7. What are the main biological themes in your target results?

Exact GO terms are often redundant because closely related concepts can appear as separate enrichment rows. RRvGO is used here only after the target-supported evidence set has been fixed. It groups Target only and Target + context terms by GO semantic similarity to provide a more readable view of the biological themes already anchored in your experiment.

```{r rrvgo-target, results='asis'}
if (!run_rrvgo) {
  cat("<div class='method-note'><strong>Semantic reduction</strong>RRvGO was not enabled for this run.</div>")
} else {
  cat(semantic_badge(semantic_orgdb, semantic_role, semantic_method))
  cat(callout(
    "how-to-read", "How to read these figures",
    "RRvGO groups related GO terms to reduce redundancy. It does not add statistical evidence, combine p-values or change which terms are supported by the target analysis."
  ))
  target_files <- rrvgo_cluster_files[grepl("target_supported", rrvgo_cluster_files, ignore.case = TRUE)]
  for (ont in c("BP", "MF", "CC")) {
    st <- rrvgo_status_for("target_supported", ont)
    path <- target_files[grepl(paste0("rrvgo_", ont, "_clusters\\.csv$"), target_files, ignore.case = TRUE)]
    status <- if (nrow(st)) as.character(st$status[[1]]) else if (length(path)) "GENERATED" else "SKIPPED_TOO_FEW_TERMS"
    if (identical(status, "GENERATED") && length(path)) {
      path <- path[[1]]
      ont <- sub("^.*rrvgo_([A-Za-z]+)_clusters\\.csv$", "\\1", path)
      root <- dirname(path)
      primary <- find_first(c(
        file.path(root, paste0("rrvgo_", ont, "_treemap.png")),
        file.path(root, paste0("rrvgo_", ont, "_bubbleplot.png"))
      ))
      if (!is.na(primary)) {
        side <- plot_sidecars(primary)
        side$png <- primary
        cat(figure_card(
          paste0("Target-supported semantic themes - ", ont),
          side$png,
          "Related target-supported exact GO terms are grouped by semantic similarity. The complete exact-term evidence remains unchanged in the Evidence Landscape.",
          report_dir, base_dir,
          side$pdf, side$svg
        ))
      } else {
        pdf_primary <- find_first(c(
          file.path(root, paste0("rrvgo_", ont, "_treemap.pdf")),
          file.path(root, paste0("rrvgo_", ont, "_bubbleplot.pdf"))
        ))
        if (!is.na(pdf_primary)) cat(EchoGO::echogo_embed_pdf_toggle(paste0("Target-supported semantic themes - ", ont), pdf_primary, report_dir, base_dir))
      }
    } else if (grepl("^SKIPPED_", status)) {
      reason <- if (nrow(st) && "reason" %in% names(st)) st$reason[[1]] else NA_character_
      cat(paste0("<div class='method-note'><strong>", ont, " semantic map</strong>", rrvgo_status_message(st), if (!is.na(reason) && nzchar(reason)) paste0(" ", reason) else "", "</div>"))
    } else if (grepl("^ERROR_", status)) {
      stop("RRvGO report validation failed for target-supported ", ont, ": ", rrvgo_status_message(st), call. = FALSE)
    }
  }
  cat(callout(
    "meaning", "What this means",
    "Use this semantic view to see the broad biological structure already supported by the target analysis. Return to the Evidence Landscape when you need the exact context-by-context evidence for a particular GO term."
  ))
  proxy_text <- if (identical(semantic_role, "proxy")) {
    paste0("RRvGO requires a GO database to calculate similarity among GO terms. <strong>", semantic_orgdb, "</strong> is used here only as a proxy for this semantic grouping step. It is not an enrichment reference and does not contribute additional enrichment evidence.")
  } else {
    paste0("<strong>", semantic_orgdb, "</strong> is used only to calculate semantic similarity among the already selected GO terms. It does not add enrichment evidence.")
  }
  cat(callout("caution", "Keep in mind", proxy_text))
}
```

# 8. What new biological themes do the annotation contexts suggest?

Context-derived hypotheses are reduced separately from the target-supported terms. This prevents semantic similarity from blending what your target analysis already supports with the hypothesis layer.

```{r rrvgo-hypothesis, results='asis'}
if (run_rrvgo) {
  cat(semantic_badge(semantic_orgdb, semantic_role, semantic_method))
  cat(callout(
    "how-to-read", "How to read these figures",
    "These themes organize related context-derived hypotheses for follow-up. A coherent semantic theme can point to an interesting biological question, but it does not promote those terms to target-supported findings."
  ))
  hyp_files <- rrvgo_cluster_files[grepl("alternative_context_hypothesis", rrvgo_cluster_files, ignore.case = TRUE)]
  for (ont in c("BP", "MF", "CC")) {
    st <- rrvgo_status_for("alternative_context_hypothesis", ont)
    path <- hyp_files[grepl(paste0("rrvgo_", ont, "_clusters\\.csv$"), hyp_files, ignore.case = TRUE)]
    status <- if (nrow(st)) as.character(st$status[[1]]) else if (length(path)) "GENERATED" else "SKIPPED_TOO_FEW_TERMS"
    if (identical(status, "GENERATED") && length(path)) {
      path <- path[[1]]
      ont <- sub("^.*rrvgo_([A-Za-z]+)_clusters\\.csv$", "\\1", path)
      root <- dirname(path)
      primary <- find_first(c(
        file.path(root, paste0("rrvgo_", ont, "_treemap.png")),
        file.path(root, paste0("rrvgo_", ont, "_bubbleplot.png"))
      ))
      if (!is.na(primary)) {
        side <- plot_sidecars(primary)
        side$png <- primary
        cat(figure_card(
          paste0("New biological themes from alternative contexts - ", ont),
          side$png,
          "Related context-derived hypotheses are grouped by semantic similarity. They remain separate from the biological themes already anchored in your target analysis.",
          report_dir, base_dir,
          side$pdf, side$svg
        ))
      } else {
        pdf_primary <- find_first(c(
          file.path(root, paste0("rrvgo_", ont, "_treemap.pdf")),
          file.path(root, paste0("rrvgo_", ont, "_bubbleplot.pdf"))
        ))
        if (!is.na(pdf_primary)) cat(EchoGO::echogo_embed_pdf_toggle(paste0("New biological themes from alternative contexts - ", ont), pdf_primary, report_dir, base_dir))
      }
    } else if (grepl("^SKIPPED_", status)) {
      reason <- if (nrow(st) && "reason" %in% names(st)) st$reason[[1]] else NA_character_
      cat(paste0("<div class='method-note'><strong>", ont, " hypothesis semantic map</strong>", rrvgo_status_message(st), if (!is.na(reason) && nzchar(reason)) paste0(" ", reason) else "", "</div>"))
    } else if (grepl("^ERROR_", status)) {
      stop("RRvGO report validation failed for alternative-context ", ont, ": ", rrvgo_status_message(st), call. = FALSE)
    }
  }
  cat(callout(
    "meaning", "What this means",
    paste0(
      "The semantic map helps move from a long list of new exact terms to a smaller set of biological questions worth examining. Examples among the first displayed context-derived terms include: <em>",
      format_examples(hypothesis_examples), "</em>."
    )
  ))
  cat(callout(
    "caution", "Keep in mind",
    "These are still hypotheses from alternative contexts. Their value is that they are traceable to exact GO terms, exact annotation contexts and contributing genes - not that they receive a higher combined score."
  ))
}
```

# 9. Which functions are connected through the same genes?

The network view answers a different question from RRvGO. RRvGO groups GO terms because their meanings are similar in the ontology. These networks connect GO terms when they share contributing genes from your experiment.

```{r network-how, results='asis'}
cat(callout(
  "how-to-read", "How to read these networks",
  "Node size reflects contributing-gene count. Node colour marks a graph module. In the target-supported network, node shape distinguishes target-only terms from target findings also recovered in alternative contexts. Edges join terms that share contributing genes. These are descriptive relationships, not combined significance."
))
```

## Target-supported gene-overlap network

This network contains only GO terms supported by your target analysis. It shows how that biological interpretation is connected through shared contributing genes.

```{r target-networks, results='asis'}
render_network_product <- function(product, ont, title, caption) {
  row <- if (is.data.frame(network_status) && nrow(network_status)) {
    dplyr::filter(network_status, .data$semantic_product == .env$product, toupper(.data$ontology) == toupper(.env$ont))
  } else tibble::tibble()
  if (!nrow(row)) {
    cat(paste0("<div class='method-note'><strong>", title, "</strong>No network product status was recorded.</div>"))
    return(invisible(NULL))
  }
  status <- as.character(row$status[[1]])
  if (identical(status, "ERROR_RENDERING")) stop(paste0(title, ": ", row$status_message[[1]]), call. = FALSE)
  if (!identical(status, "GENERATED")) {
    cat(paste0("<div class='method-note'><strong>", title, "</strong>", row$status_message[[1]], "</div>"))
    return(invisible(NULL))
  }
  png <- row$png_file[[1]]
  if (is.na(png) || !file.exists(png)) stop(paste0(title, " is marked GENERATED but has no PNG."), call. = FALSE)
  cat(figure_card(title, png, caption, report_dir, base_dir, row$pdf_file[[1]], row$svg_file[[1]]))
  links <- character()
  for (spec in list(c("GraphML", "graphml_file"), c("Nodes", "nodes_file"), c("Edges", "edges_file"), c("Interactive HTML", "html_file"))) {
    path <- row[[spec[[2]]]][[1]]
    if (!is.na(path) && nzchar(path) && file.exists(path)) {
      u <- stage_asset(path, report_dir, base_dir)
      if (!is.na(u)) links <- c(links, paste0("<a href='", u, "' target='_blank' rel='noopener'>", spec[[1]], "</a>"))
    }
  }
  if (length(links)) cat(paste0("<div class='download-bar'><strong>Network products:</strong> ", paste(links, collapse = " "), "</div>"))
}
for (ont in c("BP", "MF", "CC")) render_network_product("target_supported", ont, paste0("Target-supported gene-overlap network - ", ont), "Only target-supported terms are shown. Edges indicate shared contributing genes and are separate from RRvGO semantic similarity.")
```

## New-hypothesis gene-overlap network

This network contains only the GO terms added by selected alternative annotation contexts. It asks whether those additional hypotheses are being driven by overlapping sets of genes.

```{r hypothesis-networks, results='asis'}
for (ont in c("BP", "MF", "CC")) render_network_product("alternative_context_hypothesis", ont, paste0("Context-derived hypothesis gene-overlap network - ", ont), "Only terms from the context-derived hypothesis layer are shown. Dense local modules can indicate different GO descriptions being driven by overlapping parts of the same gene-level response.")
cat(callout(
  "meaning", "What this means",
  "The target-supported and new-hypothesis networks remain separate so that shared genes cannot blur the boundary between what your target analysis already supports and what the alternative contexts merely suggest for follow-up."
))
cat(callout(
  "caution", "Keep in mind",
  "Network connectivity depends on the available contributing-gene labels and the shared-gene threshold. A disconnected or absent term is not evidence that a biological process is unimportant."
))
```

# 10. Why did this new hypothesis appear?

A useful hypothesis should be inspectable. This section follows one context-derived term from its EchoGO classification back to the exact contexts in which it qualified, the source-local enrichment rows and the contributing genes reported by those sources.

```{r drilldown, results='asis'}
hypotheses <- if (nrow(evidence)) evidence |>
  filter(.data$evidence_profile == "ALTERNATIVE_CONTEXT") |>
  arrange(.data$display_order, .data$term_id) else tibble::tibble()

selected_id <- params$drilldown_term_id %||% if (nrow(hypotheses)) hypotheses$term_id[[1]] else NULL
if (is.null(selected_id) || !nzchar(selected_id)) {
  cat("<em>No context-derived hypothesis is available for drill-down in this run.</em>")
} else {
  drilldown <- tryCatch(
    EchoGO::echogo_hypothesis_drilldown(evidence, sources, selected_id, annotation),
    error = function(e) NULL
  )
  if (is.null(drilldown)) {
    cat("<em>The selected hypothesis could not be reconstructed from the current provenance tables.</em>")
  } else {
    trace_asset <- safe_save(plot_trace(drilldown), "hypothesis_evidence_trace", 12, 9)
    cat(callout(
      "how-to-read", "How to read this trace",
      "The diagram is a provenance path, not a causal pathway. It shows why the term entered the context-derived hypothesis layer and where you can inspect the supporting contexts and genes."
    ))
    cat(figure_card(
      "Why this new hypothesis appeared",
      trace_asset$png,
      "The target result, selected alternative annotation contexts and contributing genes are kept as separate parts of the evidence trail.",
      report_dir, base_dir,
      trace_asset$pdf, trace_asset$svg
    ))
    term <- drilldown$term
    cat(paste0("### ", term$term_id[[1]], ": ", term$term_name[[1]], "\n\n"))
    cat(paste0(
      "This term is shown as a **new hypothesis from alternative annotation contexts** because it lacks qualifying target GOseq support for the exact GO term but is recovered in **",
      term$alternative_context_support_n[[1]], "/", term$alternative_queried_context_n[[1]],
      "** selected alternative annotation contexts.\n\n"
    ))
    cat(table_details("Exact EchoGO evidence for this term", drilldown$term, NULL, preview_rows = params$top_n))
    cat(table_details("Source-level evidence for this term", drilldown$sources, NULL, preview_rows = params$top_n))
    if (!is.null(drilldown$annotation) && nrow(drilldown$annotation)) {
      cat(table_details("Gene-level annotation provenance", drilldown$annotation, NULL, preview_rows = params$top_n))
    }
    cat(callout(
      "meaning", "What this means",
      "This is where an EchoGO hypothesis becomes actionable: you can see which contexts contributed it, which genes connect the query to the term, and whether those gene identities and annotations are biologically credible enough to justify follow-up."
    ))
    cat(callout(
      "caution", "Keep in mind",
      "EchoGO does not validate the transferred functional interpretation. The drill-down makes the provenance visible so that you can evaluate it rather than accepting an opaque combined result."
    ))
  }
}
```

# 11. Additional exploratory results

```{r exploration, results='asis'}
if (!run_exploratory) {
  cat("<div class='exploratory-banner'><strong>Default-domain exploration was not enabled.</strong><br/>All primary EchoGO evidence shown above uses the experiment-matched background.</div>")
} else {
  cat("<div class='exploratory-banner'><strong>Exploratory layer - outside primary EchoGO evidence.</strong><br/>These g:Profiler results use each selected context database's own annotated statistical domain instead of the experiment-matched background. They are deliberately excluded from the primary evidence profiles, recurrence, primary networks and both RRvGO products.</div>")
  cat(callout(
    "how-to-read", "How to read these figures",
    "Treat these as database-wide exploratory enrichment. Keep each adjusted p-value and fold-enrichment value within the annotation contexts that generated it and do not mix these terms into the primary Evidence Landscape."
  ))
  if (nrow(contexts)) {
    for (i in seq_len(nrow(contexts))) {
      label <- contexts$context_label[[i]]
      inner <- ""
      for (ont in c("BP", "MF", "CC")) {
        gp_status <- gprofiler_report_status(sources, label, ont, exploratory = TRUE)
        p <- find_gp_plot(label, ont, exploratory = TRUE)
        if (identical(gp_status$status[[1]], "EXPLORATORY_PLOT_EXPECTED") && is.na(p)) {
          stop(
            paste0("Exploratory g:Profiler report asset missing for ", label, " ", ont,
                   " although default-domain exploratory result rows were recorded."),
            call. = FALSE
          )
        }
        side <- plot_sidecars(p)
        if (!is.na(p) && grepl("\\.png$", p, ignore.case = TRUE)) side$png <- p
        if (gp_status$status[[1]] %in% c("NO_SOURCE_ROWS", "NO_EXPLORATORY_RESULTS")) {
          inner <- paste0(
            inner,
            "<div class='empty-state'><strong>", gp_status$status[[1]],
            "</strong><br/>", gp_status$message[[1]], "</div>"
          )
        } else {
          inner <- paste0(
            inner,
            figure_card(
              paste0(label, " - exploratory database-wide g:Profiler - ", ont),
              side$png,
              "This exploratory view uses the selected context database's annotated domain rather than the experiment-matched background and is not part of primary EchoGO evidence.",
              report_dir, base_dir,
              side$pdf, side$svg
            )
          )
        }
      }
      cat(paste0("<details><summary>", label, " - exploratory plots</summary>", inner, "</details>"))
    }
  }
  cat(callout(
    "caution", "Keep in mind",
    "A term appearing only in this exploratory section should not be described as primary EchoGO support or counted in recovery across the selected alternative annotation contexts."
  ))
}
```

# 12. Technical details and downloads

This final section is for auditing and reproducibility. The biological interpretation is intentionally presented first; the complete technical evidence remains available here.

```{r technical, results='asis'}
run_cfg <- tibble::tibble(
  setting = c(
    "EchoGO version",
    "Target g:Profiler context",
    "Selected g:Profiler contexts",
    "Alternative annotation contexts",
    "Default-domain exploration",
    "RRvGO semantic reduction",
    "Semantic reference OrgDb",
    "Semantic reference role",
    "Semantic method"
  ),
  value = c(
    as.character(utils::packageVersion("EchoGO")),
    target_context_label,
    as.character(n_contexts),
    as.character(n_alt_contexts),
    if (run_exploratory) "enabled" else "disabled",
    if (run_rrvgo) "enabled" else "disabled",
    semantic_orgdb %||% "not configured",
    semantic_role %||% "not configured",
    if (run_rrvgo) semantic_method else "not run"
  )
)
print(knitr::kable(run_cfg, format = "html", caption = "Run configuration"))

if (nrow(contexts)) cat(table_details("Annotation-context configuration", contexts, NULL, preview_rows = params$top_n))
if (nrow(sources)) {
  source_summary <- sources |>
    count(.data$source_type, .data$context_label, .data$context_role, .data$background_mode, .data$domain_scope, name = "term_rows") |>
    arrange(.data$source_type, .data$context_label)
  cat(table_details("Source and annotation-context row counts", source_summary, NULL, preview_rows = params$top_n))
}
cat(table_details("Complete exact-term evidence", evidence, evidence_url, preview_rows = params$top_n))
cat(table_details("Complete source provenance", sources, sources_url, preview_rows = params$top_n))
cat(table_details("Mapping audit", mapping, mapping_url, preview_rows = params$top_n))
if (nrow(annotation)) cat(table_details("Annotation provenance", annotation, NULL, preview_rows = params$top_n))
```

<div class="report-scope"><strong>Where to continue</strong>Use <code>consensus/term_evidence_exact.csv</code> for the complete scoreless evidence table, <code>consensus/term_source_provenance_long.csv</code> for source-level statistics and context provenance, <code>gprofiler/submitted_vectors/mapping_table.csv</code> for identifier representability, <code>rrvgo/</code> for semantic summaries when enabled, and <code>networks/</code> or the network figures above for gene-overlap interpretation.</div>
